Autonomous Oceanographic Sampling Designs using Excursion Sets for Multivariate Gaussian random fields

### 2020-06-04 main_new.tex
updates

  Improving and optimizing oceanographic sampling is a crucial task for marine science and maritime management. Faced with limited resources to understand processes in the water-column, the combination of statistics and autonomous robotics provides new opportunities for experimental designs. In this work we develop methods for efficient spatial sampling applied to the mapping of coastal processes by providing informative descriptions of spatial characteristics of ocean phenomena. Specifically, we define a design criterion based on improved characterization of the uncertainty in the excursions of vector-valued Gaussian random fields, and derive tractable expressions for the expected Bernoulli variance reduction in such a framework. We demonstrate how this criterion can be used to prioritize sampling efforts at locations that are ambiguous, making exploration more effective. We use simulations to study the properties of methods and to compare them with state-of-the-art approaches, followed by results from field deployments with an autonomous underwater vehicle as part of a case study mapping the boundary of a river plume. The results demonstrate the potential of combining statistical methods and robotic platforms to effectively inform and execute data-driven environmental sampling.

@@

\begin{keyword}
\kwd{Ocean Sampling}
\kwd{Excursion Sets}
\kwd{Gaussian Processes}
\kwd{Experimental Design}
\kwd{Adaptive Information Gathering}
\end{keyword}

@@

Monitoring the world's oceans has gained increased importance in 
light of the changing climate and increasing anthropogenic impact. A
central problem for understanding these factors is the lack of
representative data with sufficient resolution. Most of this
\emph{undersampling} can be attributed to the large spatio-temporal
variations on which ocean processes transpire, prompting the need for
effective means of sampling. By \emph{sampling}, we refer to the design of observational strategies in the spatial domain, where the use autonomous robotic platforms can be combined with statistical methods to pursue measurements with high scientific relevance. This combination, of statistical tools and robotic platforms, constitute a methodological basis for addressing the problem of efficient sampling of the ocean, which is at the core of this the paper.

There are several obstacles when doing sampling with autonomous platforms in the ocean. First, full numerical ocean models based on complex differential equations cannot provide accurate results for online operations because the onboard computers lack the computational power required. Hence statistical proxy models of the environment must be used. For our purpose, we rely on Gaussian process (GP) representations of the ocean variables of interest. Second, with undersampling issues one must usually plan for active learning during the operation, where a relevant criterion is used to extract the most valuable designs. The acquired in-situ information must then be assimilated into the statistical model onboard the platform, so that it can be used to re-compute the sampling criterion and in this way inform decisions on where to sample next. We use a criterion focusing on the prediction of excursion sets (ESs) for the multivariate spatial processes of interest. Third, the number of sampling designs is enormous, creating a trade-off between optimization and computability. One must often compromise in both development and practical operations, and we will rely on short-horizon optimization routines for sequential designs among nearest neighbors points in a spatial grid (waypoint graph).

The imperfect understanding of causality coupled with the underlying fundamental uncertainty of the ocean implies that we need to obtain more measurements and use statistical models and methods to determine what and how phenomena form, to increase the quality of predictive statements. With this in mind, we design AUV sampling strategies for reducing the spatially-integrated Bernoulli variance (IBV) in the ES indicator variables. Recent work on ESs and associated excursion probabilities (EPs) connected to spatial statistics include \cite{french2013spatio,bolin2015excursion,french2016credible}. Our focus is on uncertainty characterization and how this is influenced by sampling. In this sense, our work is similar to \cite{bect2012,chevalier2014fast,azzimonti2016quantifying} who describe analytical results for the IBV for univariate processes. A major contribution of our work is to derive closed-form results for this design criteria for situations where the underlying model is based on multivariate GPs. Another contribution is that we embed this approach in a sequential strategy to produce sampling algorithms, and apply this to a real-world scenario of autonomously sampling temperature and salinity variables using an autonomous underwater vehicle (AUV), with the intention of understanding the between the ocean and a river plume.

Motivating examples relevant for ESs of multivariate processes are abundant. In medicine, doctors do not rely solely on a single symptom but must see several combined effects before making a diagnosis. In our context of environmental sampling applications, the salinity and temperature excursions of a river plume can similarly help characterize the underlying oceanographic and bio-geochemical processes \citep{hopkins2013detection,Pinto2018}, where sampling should be aimed at mapping relevant parts of the domain to reduce uncertainty in the temperature and salinity ES. Unlike what has been done in previous plume exploration studies, we emphasize the links to spatial statistical modeling and targeted multivariate sampling criteria for sequential sampling.

The remainder of this paper is organized as follows: Section \ref{sec:bg} provides select background on ocean sampling; Section \ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV for vector-valued GPs; Section \ref{sec:heuristics} builds on these assumptions when deriving the sequential design criteria for adaptive sampling; Section \ref{sec:simulations} discusses properties of the methods in simulation studies. Section \ref{sec:case_study} demonstrates the methodology used in field work characterizing a river plume; and Section \ref{sec:concl_disc} contains a summary and a discussion of future work.

\section{Background on Ocean Processes and Observations}\label{sec:bg}

Central to understanding the changes taking place in the upper
water-column is knowledge of the biophysical interaction driven by
an agglomeration of physical forcings (e.g. wind, topography, bathymetry, tidal influences, etc.) and incipient micro-biology driven by planktonic and coastal anthropogenic input, such as pollution and agricultural runoff transported into the ocean by rivers and streams. These often result in a range of ecosystem-related phenomena such as blooms and plumes, with direct and indirect effects on society.

Traditional data collection to understand such phenomena at sea has typically been based on static buoys, floats, or ship-based methods, with significant logistical limitations that directly impact coverage and sampling resolution. Modern methods using satellite remote-sensing provide large-scale coverage but have limited resolution, are limited to sensing the surface, and are impacted by cloud cover. Numerical ocean models similarly find it challenging to provide detail at smaller scale \citep{Lermusiaux:2006}.  The advent of mobile robotic platforms \citep{Bellingham07} has resulted in significant contributions to environmental monitoring and sampling. In particular, AUVs have advanced the state of in-situ sampling, and made robotics an integral part of ocean observation, filling parts of the undersampling gap (See Fig. \ref{fig:envir1}).

@@

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network and the sense-plan-act control methodology.]{\includegraphics[width =
    0.49\textwidth]{Figures/envir2.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Frontal patterns off of the Nidelva river, Trondheim, Norway.]{\includegraphics[width =
    0.49\textwidth]{Figures/river_proccess.pdf}\label{fig:nidelven}}
  \caption{\ref{fig:envir1} Traditional ocean observation based on
    ship-based sampling has recently been augmented by autonomous robotic vehicles. % and their interactions. 
    AUV platforms are an integral part of this network being able to reason and make decisions for efficient onboard adaptive sampling. % using the sense-plan-act control approach to autonomous control.
    \ref{fig:nidelven} The interaction of river and ocean creates
    processes that are challenging to map, where the combination of
    statistics and robotics can play a vital role in enabling more
    effective observation practices.}
\label{fig:envir}
\end{figure}

The AUVs considered here operate at $1$-$3$ m/s in the upper water column, where they are free to move in six degrees of freedom (6 DOF). The in-water operation time capacity depends on survey speed, payload sensors and navigation. Generally this amounts to 8-24 hrs. operation for smaller crafts, and 24-48 hrs. for larger crafts. AUVs have limited computational capacity, as they typically use single-board computers (SBCs), like the Raspberry Pi or a multicore GPU NVIDIA Jetson TX1 (quad-core 1.91 GHz 64-bit ARM machine, a 2-MB L2 shared cache, and 4 GB of 1600 MHz DRAM) to do computation.

Surveys with AUVs are usually limited to observations along fixed transects that are pre-scripted in mission plans created manually by a human operator. The mission can be specified operating on a scale of hundreds of meters to tens of kilometers depending on the scientific context. Faced with limited coverage capacity, a more effective approach is to instead use onboard algorithms to continuously evaluate, update, and refine future sampling locations (sense-plan-act cycle in Fig. \ref{fig:envir1}), making the information gathering \emph{adaptive} \citep{das11b,Das2015,fossuminformation,fossum18b}. This usually occurs on a spatial grid, called a waypoint graph. In doing so, the space of sampling opportunities is still limited by the waypoint graph, but the AUV has flexibility to modify its path at each waypoint based on what has been measured onboard \citep{py10,Rajan12,Rajan12b}.

The application that is of focus in this work is towards spatial characterization of a frontal system generated by river plumes. Fig. \ref{fig:nidelven} shows the survey area (Nidelva, Norway) where cold freshwater enters from the river, creating a strong gradient in both temperature and salinity. Because of the local topography and the Coreolis force %\citep{coriolis1835memoire} 
the cold fresh water tends to flow near
land to the east. Depending on the river discharge, tidal effects, wind, and temperature differences, this boundary often gets distorted. Initial knowledge about the location and evolution of these but the locations of these features are highly uncertain, making deterministic planning challenging.

Adaptive in-situ AUV sampling of an evolving frontal feature has been explored in \cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches typically use a reactive-adaptive scheme, whereby exploration does not rely on a statistical model of the environment, but rather adapts based on closing the sensing and actuation loop. Myopic sampling, i.e. stage-wise selection of the path (on the waypoint graph), has been used for surveys \citep{singh2009efficient,Binney2013} that focus largely on reducing predictive variance or entropy. These criteria are widely adopted in the statistics literature on spatio-temporal design as well \cite{bueso1998state,zidek2019monitoring}, but variance and entropy reduction are independent of the actual data realizations under the assumptions of GP models, so limited decision flexibility is created for sequential designs. The use of data-driven adaptive criteria was introduced to include more targeted sampling of regions of scientific interest in \cite{Low2009} and \cite{fossuminformation}. In this paper, with a focus on mapping the river plume by rewarding the designs that improve the classification of ES in temperature and salinity.

@@

@@

Our objective is to leverage statistical tools onboard autonomous vehicles to characterize a river
plume, focusing on spatial separation of cold freshwater from the
river and warmer saline waters of the fjord. The characterization of uncertainties on the ES is a useful
goal of this experimental sampling.

In order to formalize the problem in mathematical terms,
let us consider a multivariate (i.e., vector-valued) Gaussian random field $(\gp[\x])_{\x \in \mathcal{M}}$, where $\mathcal{M}$ is an arbitrary set (typically in geostatistical studies, a subset of interest of the ambient Euclidean space or of a sphere therein) and the dimensionality of the output $\no \geq 1$ is arbitrary. Assume furthermore that the set of interest is characterized in terms of a subset of values of the variables, say $\T \subset \R^{\no}$, and denote the corresponding set of index points by 
$$
\es=\gp^{-1}(\T)=\{\x \in \mathcal{M}: \gp[\x] \in \T\}.
$$
Our goal here is to develop approaches to quantify and improve the characterization of uncertainties on such sets in the context of multivariate random fields.

While some aspects of the developed
approaches do not call for a specific form of $\T$, we will often stick for
simplicity to the case of orthants ($\T=(-\infty, t_1] \times \dots \times (-\infty, t_{\no}]$ where $t_1,\dots, t_{\no} \in \R$) as this will allow efficiently calculating several key quantities. Note that changing some $\leq$ inequalities to $\geq$ ones would lead to immediate adaptations. In the application specific properties following the orthant assumption come into play: We let $t_1$ be a threshold for temperature and $t_2$ a threshold for salinity. The set of index point then defines $\mbox{ES} = \{\x \in D : \gp[\x,1] \leq t_1,\gp[\x,2]  \leq t_2\}$. The associated EPs are $p(\x) = P(\gp[\x,1] \leq t_1,\gp[\x,2]  \leq t_2)$, $\x \in \mathcal{M}$.

For statistical generality, we will however keep the abstract notation with $\T$ and general $\es$ for most of the derivations. Note that assuming that $\gp$ has continuous trajectories (almost surely) and $T$ to be closed, then $\es$ can be cast as a Random Closed Set \citep{Molchanov2005,adler2009random}. %In what follows, we will express various quantities related to $\es$ in terms of $\gp$'s moments both for general and orthant $T$'s. 
Let us in particular consider a locally finite (Borel) measure $\mes$ on $D$ and investigate the probability distribution of $\mes(\es)$ through its moments.

@@

	where the product measure is denoted as
	$\nu^{\otimes}:=\bigotimes_{i=1}^r \nu$. 
	%Note that we will use boldface to denote concatenated variables. % TO RELOCATE
	Here $\gp$ is defined on $\mathcal{M}$, and for
	$\bm{u}=\left(u^{(1)}, ..., u^{(r)}\right)\in \mathcal{M}^r$, $\gp[\bm{u}]=\left(\gp[u^{(1)}], ...,
	\gp[u^{(r)}]\right)\in \mathbb{R}^{\no r}$.
	\medskip

	In the particular case where $\gp$ is a multivariate Gaussian random field 
	%and denoting the mean and covariance matrix of $\gp[\bm{u}]$ by $\meanUU\in\mathbb{R}^{p\times r}$ and $\covUU$, respectively, 
	we have 
	\begin{align*}
	\jointExcuProb = \mathcal{N}_{\no r}(T^r; \meanUU, \covUU),
	\end{align*}
	where $\mathcal{N}_{\no r}(\cdot ; \meanUU, \covUU)$ is the Gaussian measure on $\mathbb{R}^{\no r}$ with mean $\meanUU$ and covariance matrix $\covUU$, respectively defined blockwise by  
	%where both are defined blockwise by 
	\begin{align*}
	\meanUU&=\begin{pmatrix}\mu(u^{(1)})\\ \vdots\\ \mu(u^{(r)})\end{pmatrix}
	\in \mathbb{R}^{\no r}, \\
	\text{and } \covUU &= \begin{pmatrix}
	\cov(\gp[u^{(1)}], \gp[u^{(1)}]) & \dots & \cov(\gp[u^{(1)}],
	\gp[u^{(r)}])\\
	\vdots & & \vdots\\
	\cov(\gp[u^{(r)}], \gp[u^{(1)}]) & \dots & \cov(\gp[u^{(r)}],
	\gp[u^{(r)}])\\
	\end{pmatrix}\in \mathbb{R}^{pr\times pr},
	\end{align*}
	each of the $r\times r$ blocks of the latter matrix being itself a (cross-)covariance matrix of dimension $\no \times \no$. 
	%dimensional Gaussian random vector. 
	Assuming further that $\covUU$ is non-singular, the probability of interest can be formulated in terms of the $\no r$-dimensional Gaussian probability density function 
	$\varphi_{\no  r}(\cdot;~\meanUU, \covUU)$ as 
	%$\varphi_{\no \times r}$ as
	%$\bm{u}\in D^r$ such that $\gp[\bm{u}]$ be non-degenerate, the integrand can be expressed in terms of the $r \times \no$-dimensional Gaussian distribution, namely 
	%cumulative distribution function, namely 
	\begin{equation*}
	\begin{split}
	&\jointExcuProb
	=
	\int_{T^r} \varphi_{\no  r}\left(\bm{v};
	~\meanUU, \covUU\right) 
	\mathrm{d}\bm{v},
	\end{split}
	\end{equation*}
	In the particular orthant case with $\T=(-\infty, t_1] \times \dots \times (-\infty, t_{r}]$,
	the latter probability directly writes in terms of the multivariate Gaussian
	cumulative distribution, % with mean $\meanUU$ and covariance matrix $\covUU$,
	this time by the way without requiring $\covUU$ %the latter 
	to be non-singular: 
	\begin{equation*}
	\begin{split}
	\jointExcuProb 
	&=
	\varPhi_{\no r}\left(\bm{t};~\meanUU, \covUU\right),
	\end{split}
	\end{equation*}
	where we have used the notations 
	$t=(t_1,\dots,t_{\no}%,..., t_1, ..., t_p
	)\in\mathbb{R}^{\no}$, $1_{r}=(1,\dots,1)\in \R^{r}$, and 
	$\bm{t}=1_{r}\otimes \bm{t}=(t_1,\dots,t_{\no},\dots,t_1,\dots,t_{\no}) 
	\in \R^{\no r}$.
\end{propo}

>>>>

  Improving and optimizing oceanographic sampling is a crucial task
  for marine science and maritime management. Faced with limited
  resources to understand processes in the water-column, the
  combination of statistics and autonomous robotics provides new
  opportunities for experimental design. In this work we develop
  methods for efficient spatial sampling applied to the mapping of
  coastal ocean processes by providing informative descriptions of
  spatial characteristics of ocean phenomena. Specifically, we define
  a design
  criterion based on improved characterization of the uncertainty in
  the excursions of vector-valued Gaussian random fields, and derive
  tractable expressions for the expected Bernoulli variance reduction
  in such a framework. We demonstrate how this criterion can be used
  to prioritize sampling efforts at locations that are ambiguous,
  making exploration more effective. We use simulations to study the
  properties of methods and to compare them with state-of-the-art
  approaches, followed by results from field deployments with an
  autonomous underwater vehicle (AUV) as part of a study mapping the
  boundary of a river plume. The results demonstrate the potential of
  combining statistical methods and robotic platforms to effectively
  inform and execute data-driven environmental sampling.

@@

\begin{keyword}
\kwd{Ocean Sampling}
\kwd{Excursion Sets}
\kwd{Gaussian Processes}
\kwd{Experimental Design}
\kwd{Autonomous robots}
\kwd{Adaptive Information Gathering}
\end{keyword}

@@

Monitoring the world's oceans has gained increased importance in light
of the changing climate and increasing anthropogenic impact. A central
problem for understanding these factors is the lack of representative
data with sufficient resolution. Most of this \emph{undersampling} can
be attributed to the large spatio-temporal variations in which ocean
processes transpire, prompting the need for effective means of
sampling. By \emph{sampling}, we refer to the design of observational
strategies in the spatial domain, where the use autonomous robotic
platforms can be combined with statistical methods to pursue
measurements with high scientific relevance. This combination, of
statistical tools and robotic platforms, constitute a methodological
basis for addressing the problem of efficient sampling of the ocean,
which is at the core of this work.

There are several challenges with sampling with autonomous platforms.
\kc{You seem to jump the gun in stating a challenge without
  connecting it to the sampling problem. Introduce a sentence as to
  why models (as a proxy) are important for \emph{understanding} ocean
  processes} First, full numerical ocean models based on complex
differential equations cannot be run onboard a marine robot with
limited computational capacity.
% provide accurate results for online
% operations because the onboard computers lack the computational power
% required.
Hence statistical proxy models of the environment must be
used. For our purpose, we rely on Gaussian process (GP)
representations of the ocean variables of interest.\kc{Sorry....but
  this makes no sense whatsoever. What is the connection of you using
  a GP with sampling? You need to make a case first as to the
  viability of using GP in a marine robot first as a potential viable
  option, before you state why you want to use a GP}. Second, with
undersampling issues \kc{what issues? If there are issues, you state
  them first and then suggest solutions.} one must usually plan for
active learning during the operation, where a relevant criterion is
used to extract the most valuable designs. The acquired in-situ
information must then be assimilated into the statistical model
onboard the platform, so that it can be used to re-compute the
sampling criterion and in this way inform decisions on where to sample
next. We use a criterion focusing on the prediction of excursion sets
(ESs) for the multivariate spatial processes of interest. Third, the
number of sampling designs is enormous, creating a trade-off between
optimization and computability. One must often compromise in both
development and practical operations, and we will rely on
short-horizon optimization routines for sequential designs among
nearest neighbors points in a spatial grid (waypoint graph).

\kc{The para above needs a rewrite to me. It starts off by stating
  there are challenges. But then jumps to models without stating the
  connection with sampling, and then jumps into learning without
  stating why learning is important. Typically the English in formal
  text starts with a statement which is supported or clarified. In the
  two cases above, the connections are simply not obvious. If this was
  there in the first version of the m/s submitted, I'm not sure why
  the reviewers didn't catch it. This is not just a stylistic issue;
  it has to do with making sense and bring the reader/reviewer along.}

\kc{Second, I don't see a motivation for GPs here and as I read
  further into the text. Even for stats folks, are GPs the ``best''
  ways to go? Are they obviously the tools one would use in
  environmental sampling? And even if they are, I think we ought to
  make the case for it strongly and generally, so this paper is
  actually accessible and understandable to those not in the Spatial
  Stats field. I would suggest adding a short para on GPs and why
  they're important -- i.e. motivate the use of GPs early, so we can
  freely make attributions to them later, as in the next section, for
  example.}

The imperfect understanding of causality coupled with the underlying
fundamental uncertainty of the ocean implies that we need to make a
large number of measurements. In addition, the use statistical models
and methods to determine what and how phenomena form, is one way to
increase the quality of predictive power aiding our understanding of
upper water-column processes.

With this in mind, we have designed AUV sampling strategies for
reducing the spatially-integrated Bernoulli variance (IBV) in the ES
indicator variables. Recent work on ESs and associated excursion
probabilities (EPs) connected to spatial statistics include
\cite{french2013spatio,bolin2015excursion,french2016credible}. Our
focus is on uncertainty characterization and how this is influenced by
sampling. In this sense, our work is similar to
\cite{bect2012,chevalier2014fast,azzimonti2016quantifying} who
describe analytical results for the IBV for univariate processes.

A major contribution of our work is to derive closed-form results for
this design criteria for situations where the underlying model is
based on multivariate GPs. Another contribution is that we embed this
approach in a sequential strategy to produce sampling algorithms, and
apply this to a real-world scenario of autonomously sampling
temperature and salinity gradients using an AUV, with the intention of
characterizing the variability of oceanic and riverine waters in
typical mixed coastal environments.

Motivating examples relevant for ESs of multivariate processes are
abundant. In medicine, doctors do not rely solely on a single symptom
but must see several combined effects before making a diagnosis. In
our context of environmental sampling, the salinity and temperature
excursions of a river plume can similarly help characterize the
underlying bio-geochemical processes
\citep{hopkins2013detection,Pinto2018}, where sampling can be aimed
towards reducing uncertainty in the temperature and salinity ES.
Unlike what has been done in previous plume exploration studies, we
emphasize the links to spatial statistical modeling and targeted
multivariate sampling criteria for sequential sampling.

The remainder of this paper is organized as follows: Section
\ref{sec:bg} provides select background on ocean sampling. Section
\ref{sec:ESEP} defines ESs, EPs, and the design criteria of IBV for
vector-valued GPs. Section \ref{sec:heuristics} builds on these
assumptions when deriving the sequential design criteria for adaptive
sampling. Section \ref{sec:simulations} discusses properties of the
methods in simulation studies. Section \ref{sec:case_study}
demonstrates the methodology used in field work characterizing a river
plume and finally, Section \ref{sec:concl_disc} contains a summary and
a discussion of future work.

\section{Ocean Processes and Observation}
\label{sec:bg}

Central to understanding the changes taking place in the upper
water-column is knowledge of the bio-geophysical interaction driven by
an agglomeration of physical forcings (e.g. wind, topography,
bathymetry, tidal influences, etc.) and incipient micro-biology driven
by planktonic and coastal anthropogenic input, such as pollution and
agricultural runoff transported into the ocean by rivers and streams.
These often result in a range of ecosystem-related phenomena such as
blooms and plumes, with direct and indirect effects on society. \kc{we
  should look for a citation here. Check with John Ryan}

Traditional data collection to understand such phenomena at sea has
typically been based on static buoys, floats, or ship-based methods,
with significant logistical limitations that directly impact coverage
and sampling resolution. Modern methods using satellite remote-sensing
provide large-scale coverage but have limited resolution, are limited
to sensing the surface, and are impacted by cloud cover. Numerical
ocean models similarly find it challenging to provide detail at fine
scale \citep{Lermusiaux:2006}. The advent of mobile robotic platforms
\citep{Bellingham07} has resulted in significant contributions to
environmental monitoring and sampling. In particular, AUVs have
advanced the state of in-situ sampling, and made robotics an integral
part of ocean observation, filling parts of the undersampling gap
\cite{rudnick03,rudnick18}(See Fig. \ref{fig:envir1}).

@@

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network and the sense-plan-act control methodology.]{\includegraphics[width =
    0.49\textwidth]{Figures/envir2.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Frontal patterns off of the Nidelva river, Trondheim, Norway.]{\includegraphics[width =
    0.49\textwidth]{Figures/river_proccess.pdf}\label{fig:nidelven}}
  \caption{\ref{fig:envir1} Traditional ocean observation based on 
    ship-based sampling has been augmented by autonomous
    robotic vehicles. % and their interactions.  
    AUV platforms are an integral part of this network being able to
    reason and make decisions for efficient onboard adaptive sampling.
    % using the sense-plan-act control approach to autonomous control. 
    \ref{fig:nidelven} The interaction of river and ocean creates
    processes that are challenging to map, where the combination of
    statistics and robotics can play a vital role in enabling more
    effective oceanographic observation.\kc{The inset ``AUV robot''
      looks very tacky and ad-hoc; might be best to remove it. Why not
      add a subfigure of an AUV instead and not lump it into a figure
      which is already quite busy? It can't hurt. Change ``ships'' to
      ``Ship'', remove the aerial drone, since we didn't use it and
      don't plan to in this work.}}
  \label{fig:envir} \end{figure}

The AUVs considered here powered untethered platforms that operate at
$1$-$3$ m/s in the upper water column, where they are free to move in
six degrees of freedom (6 DOF). The in-water operation time capability
depends on survey speed, payload sensors and navigation; typically
this amounts to 8-48 hrs. AUVs typically use single-board computers
(SBCs), like the Raspberry Pi or a multicore GPU NVIDIA Jetson TX1
(quad-core 1.91 GHz 64-bit ARM machine, a 2-MB L2 shared cache, and 4
GB of 1600 MHz DRAM) for computation onboard.

Surveys with AUVs are usually limited to observations along fixed
transects that are pre-scripted in mission plans created manually by a
human operator. The mission can be specified operating on a scale of
hundreds of meters to tens of kilometers depending on the scientific
context. Faced with limited coverage capacity, a more effective
approach is to instead use onboard algorithms to continuously
evaluate, update, and refine future sampling locations (sense-plan-act
cycle in Fig. \ref{fig:envir1}), making the information gathering
\emph{adaptive} \citep{das11b,Das2015,fossuminformation,fossum18b}.
% This usually occurs on a spatial grid, called a waypoint graph. 
In doing so, the space of sampling opportunities is still limited by
the waypoint graph, but the AUV has flexibility to modify its path at
each waypoint based on in-situ sensing and measurements onboard
\citep{py10,Rajan12,Rajan12b}.

We focus on spatial characterization of a frontal system generated by
river plumes. Fig. \ref{fig:nidelven} shows the survey area in
Trondheim, Norway, where cold freshwater enters from the river,
creating a strong gradient in both temperature and salinity. Because
of the local topography and the Coriolis force the cold fresh water
tends to flow to the east. Depending on the river discharge, tidal
effects, wind, and temperature differences, this boundary often gets
distorted. Initial knowledge about the location and evolution of these
features are highly uncertain, making deterministic planning
challenging.

Adaptive in-situ AUV sampling of an evolving frontal feature has been
explored in \cite{fronts11,Zhang2012,Pinto2018,costa19}. These
approaches typically use a reactive-adaptive scheme, whereby
exploration does not rely on a statistical model of the environment,
but rather adaptation is based on closing the sensing and actuation
loop. Myopic sampling, i.e. stage-wise selection of the path (on the
waypoint graph), has been used for surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or entropy. These criteria are widely adopted in
the statistics literature on spatio-temporal design as well
\cite{bueso1998state,zidek2019monitoring}, but variance and entropy
reduction are independent of the actual data realizations under the
assumptions of GP models, so limited decisional flexibility is created
for sequential designs \kc{should this be, ``...so it has limited
  decisional flexibility''?}. The use of data-driven adaptive criteria
was introduced to include more targeted sampling of regions of
scientific interest in \cite{Low2009} and \cite{fossuminformation}. In
this paper, we focus on mapping the river plume by rewarding designs
that improve the classification of ES in temperature and salinity.

@@

\kc{Typically, I have been told in the past, it is bad form to start a section and
  then a subsection right away. I sort of see it, so, it might be good to
  provide a 1-2 sentence introduction here perhaps?}

@@

Our objective is to leverage statistical tools onboard an autonomous
robotic platform to characterize a river plume, focusing on spatial
separation of cold freshwater from a river and warmer saline waters
of a fjord. The characterization of uncertainties on the ES is a
useful goal of this sampling strategy.

In order to formalize the problem in mathematical terms, let us
consider a multivariate (i.e., vector-valued) Gaussian random field
$(\gp[\x])_{\x \in \mathcal{M}}$, where $\mathcal{M}$ is an arbitrary
set (typically in geostatistical studies, a subset of interest of the
ambient Euclidean space or of a sphere therein \kc{is this obvious to
  the readership why a sphere? It isn't for me certainly.}) and the
dimensionality of the output $\no \geq 1$ is arbitrary \kc{Is the
  definition of such a 'p' and its dimensionality, obvious? This to me
  springs out of nowhere.}. Assume furthermore that the set of interest
is characterized in terms of a subset of values of the variables, say
$\T \subset \R^{\no}$, and denote the corresponding set of index
points by
$$
\es=\gp^{-1}(\T)=\{\x \in \mathcal{M}: \gp[\x] \in \T\}.
$$
Our goal here is to develop approaches to quantify and improve the
characterization of uncertainties on such sets \kc{$\es$?} in the
context of multivariate random fields.

While some aspects of the developed approaches do not call for a
specific form of $\T$, we will often stick for simplicity to the case
of orthants
($\T=(-\infty, t_1] \times \dots \times (-\infty, t_{\no}]$ where
$t_1,\dots, t_{\no} \in \R$) as this will allow efficient calculation
of several key quantities. Note that changing some $\leq$ inequalities
to $\geq$ ones would lead to immediate adaptations. In the application
specific properties following the orthant assumption come into play
\kc{This last phrase doesn't parse; is it ``..., the following
  orthant....''?}: We let $t_1$ be a threshold for temperature and
$t_2$ a threshold for salinity \kc{Might be easier to remember with
  symbols 't' and 's' in some
  form.}. The set of index point then defines
$\mbox{ES} = \{\x \in D : \gp[\x,1] \leq t_1,\gp[\x,2] \leq t_2\}$.
The associated EPs are
$p(\x) = P(\gp[\x,1] \leq t_1,\gp[\x,2] \leq t_2)$,
$\x \in \mathcal{M}$.

For statistical generality, we will however keep the abstract notation
with $\T$ and general $\es$ \kc{Again, previous phrase doesn't parse.}
for most of the derivations. Note that assuming that $\gp$ has
continuous trajectories (almost surely) and $T$ to be closed, then
$\es$ can be cast as a Random Closed Set
\citep{Molchanov2005,adler2009random}. %In what follows, we will express various quantities related to $\es$ in terms of $\gp$'s moments both for general and orthant $T$'s.
Let us in particular consider a locally finite (Borel) measure $\mes$
on $D$ and investigate the probability distribution of $\mes(\es)$
through its moments.

@@

	where the product measure is denoted as
	$\nu^{\otimes}:=\bigotimes_{i=1}^r \nu$. 
	%Note that we will use boldface to denote concatenated variables. % TO RELOCATE
	Here $\gp$ is defined on $\mathcal{M}$, and for
	$\bm{u}=\left(u^{(1)}, ..., u^{(r)}\right)\in \mathcal{M}^r$, $\gp[\bm{u}]=\left(\gp[u^{(1)}], ...,
	\gp[u^{(r)}]\right)\in \mathbb{R}^{\no r}$.\kc{Is $\no r$ a
        product of the $\no$ from above? Is that obvious?}
	\medskip

	In the particular case where $\gp$ is a multivariate Gaussian random field 
	%and denoting the mean and covariance matrix of $\gp[\bm{u}]$ by $\meanUU\in\mathbb{R}^{p\times r}$ and $\covUU$, respectively, 
	we have 
	\begin{align*}
	\jointExcuProb = \mathcal{N}_{\no r}(T^r; \meanUU, \covUU),
	\end{align*}
	where $\mathcal{N}_{\no r}(\cdot ; \meanUU, \covUU)$ is the Gaussian measure on $\mathbb{R}^{\no r}$ with mean $\meanUU$ and covariance matrix $\covUU$, respectively defined blockwise by  
	%where both are defined blockwise by 
	\begin{align*}
	\meanUU&=\begin{pmatrix}\mu(u^{(1)})\\ \vdots\\ \mu(u^{(r)})\end{pmatrix}
	\in \mathbb{R}^{\no r}, \\
	\text{and } \covUU &= \begin{pmatrix}
	\cov(\gp[u^{(1)}], \gp[u^{(1)}]) & \dots & \cov(\gp[u^{(1)}],
	\gp[u^{(r)}])\\
	\vdots & & \vdots\\
	\cov(\gp[u^{(r)}], \gp[u^{(1)}]) & \dots & \cov(\gp[u^{(r)}],
	\gp[u^{(r)}])\\
	\end{pmatrix}\in \mathbb{R}^{pr\times pr},
	\end{align*}
	each of the $r\times r$ blocks of the latter matrix being itself a (cross-)covariance matrix of dimension $\no \times \no$. 
	%dimensional Gaussian random vector. 
	Assuming further that $\covUU$ is non-singular, the probability of interest can be formulated in terms of the $\no r$-dimensional Gaussian probability density function 
	$\varphi_{\no  r}(\cdot;~\meanUU, \covUU)$ as 
	%$\varphi_{\no \times r}$ as
	%$\bm{u}\in D^r$ such that $\gp[\bm{u}]$ be non-degenerate, the integrand can be expressed in terms of the $r \times \no$-dimensional Gaussian distribution, namely 
	%cumulative distribution function, namely 
	\begin{equation*}
	\begin{split}
	&\jointExcuProb
	=
	\int_{T^r} \varphi_{\no  r}\left(\bm{v};
	~\meanUU, \covUU\right) 
	\mathrm{d}\bm{v},
	\end{split}
	\end{equation*}
	In the particular orthant case with $\T=(-\infty, t_1] \times \dots \times (-\infty, t_{r}]$,
	the latter probability directly writes in terms of the multivariate Gaussian
	cumulative distribution, % with mean $\meanUU$ and covariance matrix $\covUU$,
	this time without requiring $\covUU$ %the latter 
	to be non-singular: 
	\begin{equation*}
	\begin{split}
	\jointExcuProb 
	&=
	\varPhi_{\no r}\left(\bm{t};~\meanUU, \covUU\right),
	\end{split}
	\end{equation*}
	where we have used the notations 
	$t=(t_1,\dots,t_{\no}%,..., t_1, ..., t_p
	)\in\mathbb{R}^{\no}$, $1_{r}=(1,\dots,1)\in \R^{r}$, and 
	$\bm{t}=1_{r}\otimes \bm{t}=(t_1,\dots,t_{\no},\dots,t_1,\dots,t_{\no}) 
	\in \R^{\no r}$.
\end{propo}
